Realizability-Preserving DG-IMEX Method for the Two-Moment Model of Fermion Transport \tnoteref{support}

### 2018-06-05 Documents/M1/realizableFermionicM1/sections/imex.tex
Added the definitions of IMEX coefficients.
Forgot to commit slight modification to conditions on PARSD

To prove the positivity-preserving property of the IMEX scheme, Hu et al. \cite{hu_etal_2018} rewrite the stage values in Eq.~\eqref{imexStages} in the following form
\begin{equation}
  \vect{u}^{(i)}
  =\sum_{j=0}^{i-1}c_{ij}\Big[\,\vect{u}^{(j)}+\hat{c}_{ij}\,\dt\,\vect{\cT}(\vect{u}^{(j)})\,\Big]
  +a_{ii}\,\dt\,\f{1}{\tau}\,\vect{\cQ}(\vect{u}^{(i)}),\quad i=1,\ldots,s,
  \label{eq:imexStagesRewrite}
\end{equation}
where $c_{ij}$, and $\hat{c}_{ij}=\tilde{c}_{ij}/c_{ij}$ are computed from $\tilde{a}_{ij}$ and $a_{ij}$.
(In Eq.~\eqref{eq:imexStagesRewrite}, $\vect{u}^{(0)}=\vect{u}^{n}$.)  
Two types of IMEX schemes are considered: \emph{type~A} \cite{pareschiRusso_2005} and \emph{type~ARS} \cite{ascher_etal_1997}.  
For IMEX schemes of type~A, the matrix $A$ is invertible.  
For IMEX schemes of type~ARS, the matrix $A$ can be written as
\[\left( 
   \begin{matrix} 
       0 & 0 \\ 
       0 & \hat{A}
   \end{matrix}
\right)\]
where $\hat{A}$ is invertible.
In writing the stages in the IMEX scheme in the general form given by Eq.~\eqref{eq:imexStagesRewrite}, it should be noted that $\tilde{c}_{i0}=0$ for IMEX schemes of Type~A, and $c_{i1}=\tilde{c}_{i1}=0$ for IMEX schemes of Type~ARS \cite{hu_etal_2018}.  
%The coefficients are defined as $c_{i0} = 1 - \sum_{j=1}^{i-1}c_{ij} $, 
%For $i=1,\ldots,s$, the coefficients
%\begin{align*}
%  \tilde{c}_{i0}
%  =&\left\{
%  \begin{array}{cl}
%    0  & \text{for type $A$,} \\
%   \tilde{a}_{i1} + \sum_{l=2}^{i-1}a_{il}\tilde{b}_{l1}  & \text{for type $ARS$;}
%  \end{array}
%  \right.\\
%    \tilde{c}_{i1}
%    =&\left\{
%      \begin{array}{cl}
%       \tilde{a}_{i1} + \sum_{l=2}^{i-1}a_{il}\tilde{b}_{l1} & \text{for type $A$,} \\
%       0  & \text{for type $ARS$;}
%      \end{array}
%      \right.
%\end{align*}
%and the rest coefficients were defined as in \cite{hu_etal_2018}.
For type A schemes, the positivity-preserving property follows from requiring $a_{ii}>0$ and $c_{i0}\ge0$ for $i=1,\ldots,s$, and $c_{ij},\tilde{c}_{ij}\ge0$, for $i=2,\ldots,s$, $j=1,\ldots,s-1$.  
For type ARS schemes, the positivity follows from requiring $a_{ii}>0$ and $c_{i0},\tilde{c}_{i0}\ge0$ for $i=2,\ldots,s$, and $c_{ij},\tilde{c}_{ij}\ge0$ for $i=3,\ldots,s$, $j=2,\ldots,i-1$.  
Such coefficients were given in \cite{hu_etal_2018} for GSA schemes of type A with $s=3$ and type ARS with $s=4$.  
(It was also proven that $s=3$ and $s=4$ are the necessary number of stages needed for GSA second-order positivity-preserving IMEX schemes of type A and type ARS, respectively.)

@@

Unfortunately, the correction step in Eq.~\eqref{eq:imexCorrection} deteriorates the accuracy of the IMEX scheme when applied to the moment equations in the diffusion limit ($\xi=0$, $\tau\ll 1$), where $\vect{\cK}\approx\f{1}{3}\,\cJ\,\vect{I}$.  
In the diffusion limit, to leading order in $\tau$, we then have $\vect{\cH}\approx-\f{1}{3}\,\tau\,\nabla\cJ$, which is due to a balance between the transport term and the collision term in the equation for the particle flux $\vect{\cH}$.  
The absence of the transport operator in the correction step, destroys this balance.  
We demonstrate the inferior performance of IMEX schemes with the correction step given by Eq.~\eqref{eq:imexCorrection} in the diffusion limit in Section~\ref{sec:smoothProblems}.  
We have also implemented and tested one of the IMEX schemes in Chertock et al. \cite{chertock_etal_2015}, where the transport operator is included in the correction step, and found it to perform very well in the diffusion limit.  
However, we have not been able to prove sufficiently general realizability-preserving properties with this approach, without invoking a too severe time step restriction.  
We therefore proceed to design positivity-preserving IMEX schemes, without the correction step, that perform better in the diffusion limit.  
We will limit the scope to IMEX schemes of type~ARS.

@@

  \item Positivity-preserving (conditions given by \cite{hu_etal_2018}; with $c_{\mbox{\tiny Sch}}$ maximal).
  \item Few ($\le3$) implicit stages.

>>>>

To prove the positivity-preserving property of the IMEX scheme, Hu et al. \cite{hu_etal_2018} rewrite the stage values in Eq.~\eqref{imexStages} in the following form
\begin{equation}
  \vect{u}^{(i)}
  =\sum_{j=0}^{i-1}c_{ij}\Big[\,\vect{u}^{(j)}+\hat{c}_{ij}\,\dt\,\vect{\cT}(\vect{u}^{(j)})\,\Big]
  +a_{ii}\,\dt\,\f{1}{\tau}\,\vect{\cQ}(\vect{u}^{(i)}),\quad i=1,\ldots,s,
  \label{eq:imexStagesRewrite}
\end{equation}
where $c_{ij}$, and $\hat{c}_{ij}=\tilde{c}_{ij}/c_{ij}$ are computed from $\tilde{a}_{ij}$ and $a_{ij}$.
Such that, $c_{i0} = 1 - \sum_{j=1}^{i-1}c_{ij}$, $c_{ij} = \sum_{l=j}^{i-1}a_{il}b_{lj}$, $\tilde{c}_{i0} = \tilde{a}_{i1} + \sum_{j=2}^{i-1} a_{ij}\tilde{b}_{j1}$,$\tilde{c}_{ij} = \tilde{a}_{ij} + \sum_{l=j+1}^{i-1} a_{il}\tilde{b}_{lj}$ $b_{ij} = - \frac{1}{a_{ii}} \sum_{l = j}^{i-1}a_{il}b_{lj}$, $b_{ii} = \frac{1}{a_{ii}}$, $\tilde{b}_{ij} = \frac{1}{a_{ii}}\big( -\tilde{a}_{ij} - \sum_{l = j+1}^{i-1}a_{il}\tilde{b}_{lj}\big)$. 
(In Eq.~\eqref{eq:imexStagesRewrite}, $\vect{u}^{(0)}=\vect{u}^{n}$.)  
Two types of IMEX schemes are considered: \emph{type~A} \cite{pareschiRusso_2005} and \emph{type~ARS} \cite{ascher_etal_1997}.  
For IMEX schemes of type~A, the matrix $A$ is invertible.  
For IMEX schemes of type~ARS, the matrix $A$ can be written as
\[\left( 
   \begin{matrix} 
       0 & 0 \\ 
       0 & \hat{A}
   \end{matrix}
\right)\]
where $\hat{A}$ is invertible.
In writing the stages in the IMEX scheme in the general form given by Eq.~\eqref{eq:imexStagesRewrite}, it should be noted that $\tilde{c}_{i0}=0$ for IMEX schemes of Type~A, and $c_{i1}=\tilde{c}_{i1}=0$ for IMEX schemes of Type~ARS \cite{hu_etal_2018}.  
For type A schemes, the positivity-preserving property follows from requiring $a_{ii}>0$ and $c_{i0}\ge0$ for $i=1,\ldots,s$, and $c_{ij},\tilde{c}_{ij}\ge0$, for $i=2,\ldots,s$, $j=1,\ldots,s-1$.  
For type ARS schemes, the positivity follows from requiring $a_{ii}>0$ and $c_{i0},\tilde{c}_{i0}\ge0$ for $i=2,\ldots,s$, and $c_{ij},\tilde{c}_{ij}\ge0$ for $i=3,\ldots,s$, $j=2,\ldots,i-1$.  
Such coefficients were given in \cite{hu_etal_2018} for GSA schemes of type A with $s=3$ and type ARS with $s=4$.  
(It was also proven that $s=3$ and $s=4$ are the necessary number of stages needed for GSA second-order positivity-preserving IMEX schemes of type A and type ARS, respectively.)

@@

Unfortunately, the correction step in Eq.~\eqref{eq:imexCorrection} deteriorates the accuracy of the IMEX scheme when applied to the moment equations in the diffusion limit ($\xi=0$, $\tau\ll 1$), where $\vect{\cK}\approx\f{1}{3}\,\cJ\,\vect{I}$.  
In the diffusion limit, to leading order in $\tau$, we then have $\vect{\cH}\approx-\f{1}{3}\,\tau\,\nabla\cJ$, which is due to a balance between the transport term and the collision term in the equation for the particle flux $\vect{\cH}$.  
The absence of the transport operator in the correction step, destroys this balance.  
We demonstrate the inferior performance of IMEX schemes with the correction step given by Eq.~\eqref{eq:imexCorrection} in the diffusion limit in Section~\ref{sec:smoothProblems}.  
We have also implemented and tested one of the IMEX schemes in Chertock et al. \cite{chertock_etal_2015}, where the transport operator is included in the correction step, and found it to perform very well in the diffusion limit.  
However, we have not been able to prove sufficiently general realizability-preserving properties with this approach, without invoking a too severe time step restriction.  
We therefore proceed to design positivity-preserving IMEX schemes, without the correction step, that perform better in the diffusion limit.  
We limit the scope to IMEX schemes of type~ARS.  
(It can be shown that IMEX schemes of type~A satisfying the conditions we list below Eq.~\eqref{eq:orderConditionsEx} do not exist.)

@@

  \item Positivity-preserving (conditions given by \cite{hu_etal_2018}; with $c_{\mbox{\tiny Sch}}>0$ and maximal).
  \item Few stages ($s\le3$).
